---
title: Exercise part B
description: |
  lorem ipsum dolor sit amet i do some sampling
---

Test paragraph

Hello this is my entry for exercise 1 part B i hope you enjoy

### Steppy steps

0. Set a seed so that the data is reproducible later.
```{r}
set.seed(1337)
```

1. Create a vector that contains 100 randomly sampled values from a negative binomial distribution.
```{r}
neg_bin_dist_samples <- rnbinom(n = 100, mu = 10, size = 2)
head(neg_bin_dist_samples)
```
2. Create a vector that contains log10-transformed values from the vector in step 1.
```{r}
neg_bin_dist_samples_log10 <- log(neg_bin_dist_samples+1)
head(neg_bin_dist_samples_log10)
```
3. Plot histograms of raw and log10-transformed data side by side 
```{r}
par(mfrow = c(1,2))

hist(neg_bin_dist_samples,
     main='Raw data')
hist(neg_bin_dist_samples_log10,
     main='Log10-transform data')
```
4. Display the mean and the variance of the obtained data.

* Raw data:
```{r}
mean_raw <- summary(neg_bin_dist_samples)["Mean"]
var_raw <- var(neg_bin_dist_samples)
```
<span style="color: red;">
*Therefore we obtain a mean of `r mean_raw` and a variance of `r var_raw` for our randomly sampled data.*
</span>

* Log10-transformed data:
```{r}
mean_log10 <- summary(neg_bin_dist_samples_log10)["Mean"]
var_log10 <- var(neg_bin_dist_samples_log10)
```
<span style="color: red;">
*Therefore we obtain a mean of `r mean_log10` and a variance of `r var_log10` for our log10-transformed data.*
</span>
